\thetitle

### 2020-11-08 sun_phases/spin_qudit_tomo.tex
a bit more tightening of the analytical error bound

@@

%%% text markup
\usepackage{color} % text color
\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\todo}[1]{{\color{blue} To do: #1}}

@@

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} and derive more fundamental error bounds in Appendix \ref{sec:error}.
Therein, we find that if measurement error is dominated by shot noise, then the estimate $\tilde\rho_V$ built from $n$ independent measurements of spin projection along every axis $\v v\in V$ approximates the true state $\rho$ to within a mean squared distance $\epsilon_V^2/n$, i.e.
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  \le \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the {\it error scale} $\epsilon_V^2$ is
\begin{align}
  \epsilon_V^2 \equiv
  \f1{nd} \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
  \label{eq:scale}
\end{align}
and $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell
  \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
For reference, we note that the error scale $\epsilon_V^2<\S_V^2$.

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariances between statistical errors in the transition operator basis are then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \sp{\delta_{\mu\mu'} p_\mu^{(i)} - p_\mu^{(i)} p_{\mu'}^{(i)}}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, for any multinomial probability distribution $p\equiv\p{p_1,p_2,\cdots,p_d}$ we define the function
\begin{align}
  g_\ell\p{p} \equiv \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell\mu'}
  \p{\delta_{\mu\mu'} p_\mu - p_\mu p_{\mu'}}
  = \sum_\mu t_{\ell\mu}^2 p_\mu - \p{\sum_\mu t_{\ell\mu} p_\mu}^2
  = \sigma_p^2\p{t_\ell},
\end{align}
where $t_\ell \equiv \p{t_{\ell,-s},t_{\ell,-s+1},\cdots,t_{\ell s}}$ is a vector of all $t_{\ell\mu}$, and
\begin{align}
  \sigma_p^2\p{X}
  \equiv \sum_\mu p_\mu \p{X_\mu - \sum_\nu p_\nu X_\nu}^2
  = \sum_\mu p_\mu X_\mu^2 - \p{\sum_\mu p_\mu X_\mu}^2
\end{align}
is the variance of $X$ when sampled according to the probability distribution $p$.
By expanding
\begin{align}
  t_{\ell\mu} = \sqrt{\f{2\ell+1}{d}}\, \tau_{\ell\mu},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu},
\end{align}
we can simplify
\begin{align}
  g_\ell\p{p} = \f{2\ell+1}{d} \times \sigma_p^2\p{\tau_\ell}.
\end{align}
We now seek to maximize $g_\ell\p{p}$, which will provide an upper bound for $\bbk{\epsilon_{i\ell}^2}$.
In general, variance $\sigma_p^2\p{X}$ is maximal when $p$ has equal weight on the largest and smallest values of $X$, which implies that
\begin{align}
  \sigma_p^2\p{\tau_\ell} \le \gamma_\ell^2,
  &&
  \gamma_\ell \equiv \f{\max\tau_\ell - \min\tau_\ell}{2}.
\end{align}
Altogether, using additionally the fact that $g_0\p{p}=0$, we can bound
\begin{align}
  \bbk{\epsilon_{i\ell}^2}
  = \f1n \times g_\ell\p{p^{(i)}}
  \le \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1 - \delta_{\ell,0}}.
  \label{eq:variance_bound}
\end{align}
Note that the bound in Eq.~\eqref{eq:variance_bound} is tight, as equality is achieved by the mixed state
\begin{align}
  \rho^\star
  \equiv \f12 \p{\Pi_{\v v_i,\mu_{\t{max}}} + \Pi_{\v v_i,\mu_{\t{min}}}},
\end{align}
where $\mu_{\t{max}}$ ($\mu_{\t{min}}$) is the index that maximizes (minimizes) $\tau_{\ell\mu}$.
In addition, the  Clebsch-Gordan coefficients $\tau_{\ell\mu}$ are at most $1$ in magnitude, which implies that $\gamma_\ell^2\le 1$.
However, the bound on $\gamma_\ell^2\le1$ is not tight in general, as we could have alternatively bounded
\begin{align}
  g_\ell\p{p} \le \sum_\mu t_{\ell\mu}^2 p_\mu
  \le \sum_\mu t_{\ell\mu}^2 = 1,
\end{align}
which must also hold for the probability distribution $p^\star$ that maximizes $g_\ell$, i.e.
\begin{align}
  g_\ell\p{p^\star} = \f{2\ell+1}{d} \times \gamma_\ell^2 \le 1,
  &&
  \t{so}
  &&
  \gamma_\ell^2 \le \min\Set{1, \f{d}{2\ell+1}}.
\end{align}

>>>>

\newcommand{\ZZ}{\mathbb{Z}}

@@

%%% text markup
\usepackage{color} % text color
\newcommand{\red}[1]{{\color{red} #1}}

@@

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} and derive more fundamental error bounds in Appendix \ref{sec:error}.
Therein, we find that if measurement error is dominated by shot noise, then the estimate $\tilde\rho_V$ built from $n$ independent measurements of spin projection along every axis $\v v\in V$ approximates the true state $\rho$ to within a mean squared distance $\epsilon_V^2/n$, i.e.
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  \le \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the {\it error scale} $\epsilon_V^2$ is
\begin{align}
  \epsilon_V^2 \equiv
  \f1d \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
  \label{eq:scale}
\end{align}
and $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell
  \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
For reference, we note that the error scale $\epsilon_V^2<\S_V^2$.

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariances between statistical errors in the transition operator basis are then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \sp{\delta_{\mu\mu'} p_\mu^{(i)} - p_\mu^{(i)} p_{\mu'}^{(i)}}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, for any multinomial probability distribution $p\equiv\p{p_1,p_2,\cdots,p_d}$ we define the function
\begin{align}
  g_\ell\p{p} \equiv \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell\mu'}
  \p{\delta_{\mu\mu'} p_\mu - p_\mu p_{\mu'}}
  = \sum_\mu t_{\ell\mu}^2 p_\mu - \p{\sum_\mu t_{\ell\mu} p_\mu}^2
  = \sigma_p^2\p{t_\ell},
\end{align}
where $t_\ell \equiv \p{t_{\ell,-s},t_{\ell,-s+1},\cdots,t_{\ell s}}$ is a vector of all $t_{\ell\mu}$, and
\begin{align}
  \sigma_p^2\p{X}
  \equiv \sum_\mu p_\mu \p{X_\mu - \sum_\nu p_\nu X_\nu}^2
  = \sum_\mu p_\mu X_\mu^2 - \p{\sum_\mu p_\mu X_\mu}^2
\end{align}
is the variance of $X$ when sampled according to the probability distribution $p$.
By expanding
\begin{align}
  t_{\ell\mu} = \sqrt{\f{2\ell+1}{d}}\, \tau_{\ell\mu},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu},
\end{align}
we can simplify
\begin{align}
  g_\ell\p{p} = \f{2\ell+1}{d} \times \sigma_p^2\p{\tau_\ell}.
\end{align}
We now seek to maximize $g_\ell\p{p}$, which will provide an upper bound for $\bbk{\epsilon_{i\ell}^2}$.
In general, variance $\sigma_p^2\p{X}$ is maximal when $p$ has equal weight on the largest and smallest values of $X$, which implies that
\begin{align}
  \sigma_p^2\p{\tau_\ell} \le \gamma_\ell^2,
  &&
  \gamma_\ell \equiv \f{\max\tau_\ell - \min\tau_\ell}{2}.
\end{align}
Altogether, using additionally the fact that $g_0\p{p}=0$, we can bound
\begin{align}
  \bbk{\epsilon_{i\ell}^2}
  = \f1n \times g_\ell\p{p^{(i)}}
  \le \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1 - \delta_{\ell,0}}.
  \label{eq:variance_bound}
\end{align}
Note that the bound in Eq.~\eqref{eq:variance_bound} is tight, as equality is achieved by the mixed state
\begin{align}
  \rho^\star
  \equiv \f12 \p{\Pi_{\v v_i,\mu_{\t{max}}} + \Pi_{\v v_i,\mu_{\t{min}}}},
\end{align}
where $\mu_{\t{max}}$ ($\mu_{\t{min}}$) is any index that maximizes (minimizes) $\tau_{\ell\mu}$.

Though the result in Eq.~\eqref{eq:variance_bound} is sufficient to proceed with our derivation of an improved tomographic error bound, we first take some time to better understand the bound on $\bbk{\epsilon_{i\ell}^2}$ analytically.
To this end, we first note that the Clebsch-Gordan coefficients $\tau_{\ell\mu}$ are at most $1$ in magnitude, which implies that $\gamma_\ell^2\le 1$.
Moreover, we can also bound
\begin{align}
  g_\ell\p{p} \le \sum_\mu t_{\ell\mu}^2 p_\mu
  \le \sum_\mu t_{\ell\mu}^2 = 1,
  &&
  \t{so}
  &&
  \bbk{\epsilon_{i\ell}^2} \le \f1n \times \min\Set{\f{2\ell+1}{d}, 1}.
\end{align}
If $d$ is even, then we can get a tighter analytical bound on $\bbk{\epsilon_{i\ell}^2}$ by considering
\begin{align}
  g_\ell\p{p} \le \sum_\mu t_{\ell\mu}^2 p_\mu \le t_{\ell\mu_{\t{max}}}^2,
\end{align}
where $\mu_{\t{max}}$ is any index maximizing $t_{\ell\mu}$.
The symmetries of the Clebsch-Gordan coefficients imply that $t_{\ell\mu_{\t{max}}}^2 = t_{\ell,-\mu_{\t{max}}}^2$, so if $\mu_{\t{max}} \ne0$ (which is always true when $d$ is even because $s,\mu\notin\ZZ$), then
\begin{align}
  g_\ell\p{p} \le t_{\ell\mu_{\t{max}}}^2
  = \f12 \p{t_{\ell\mu_{\t{max}}}^2 + t_{\ell,-\mu_{\t{max}}}^2}
  \le \f12 \sum_\mu t_{\ell\mu}^2
  = \f12,
  &&
  \t{so}
  &&
  \bbk{\epsilon_{i\ell}^2}
  \le \f1n \times \min\Set{\f{2\ell+1}{d}, \f12}.
\end{align}
